Finch: Sparse and Structured Array Programming with Control Flow

### 2024-04-06 sections/eval/api.tex
spelling

\subsection{Implementing Numpy's Array API in Finch}
In the past decade, the adoption of the Python Array API \cite{harris_array_2020} has allowed for a proliferation array programming systems, but existing implementations of this and similar APIs for structured data suffer from either incompleteness or inefficiency.
%
These APIs have hundreds of required functions, from mapreduce to slicing and windowing. 
%
Compilers with simpler interfaces (such as TACOs simple einsum) need a massive amount of glue code outside of the normal compilation path to support the full API, leading to a large implementation burden.
%
As the API's were designed for only one structure (Dense), the burden only grows with the introduction of as additional structures such as sparsity, as these structures interact with the complex interface and with each other.
%
%In order to reduce the complexity of interactions between different formats, they're either limited to vectors and matrices or only support tabular representations.
%
Further, implementing each operation on its own is insufficient. Operator fusion is required to avoid asymptotic performance degradation on sparse kernels, as we show in Figure~\ref{fig:fusion}.
%
A flexible compiler like Finch, which can produce efficient code for arbitrary operations between inputs in a wide variety of formats, is the missing piece needed to implement the full array API for structured data with competitive performance.
%
To handle the expansiveness of the Array API while preserving opportunities for fusion and whole workflow optimization, we pursued a lazy evaluation strategy mediated by a high-level query language.
%
This is implemented by 1) Finch Logic, a minimal, high-level language for expressing array operations and 2) the Physical Interpreter, which executes Finch Logic as a sequence of Finch programs.

>>>>

\subsection{Implementing NumPy's Array API in Finch}
In the past decade, the adoption of the Python Array API \cite{harris_array_2020} has allowed for a proliferation array programming systems, but existing implementations of this and similar APIs for structured data suffer from either incompleteness or inefficiency.
%
These APIs have hundreds of required functions, from mapreduce to slicing and windowing. 
%
Compilers with simpler interfaces (such as TACO's simple Einsum) need a massive amount of glue code outside of the normal compilation path to support the full API, leading to a large implementation burden.
%
As the API's were designed for only one structure (Dense), the burden only grows with the introduction of as additional structures such as sparsity, as these structures interact with the complex interface and with each other.
%
%In order to reduce the complexity of interactions between different formats, they're either limited to vectors and matrices or only support tabular representations.
%
Further, implementing each operation on its own is insufficient. Operator fusion is required to avoid asymptotic performance degradation on sparse kernels, as we show in Figure~\ref{fig:fusion}.
%
A flexible compiler like Finch, which can produce efficient code for arbitrary operations between inputs in a wide variety of formats, is the missing piece needed to implement the full array API for structured data with competitive performance.
%
To handle the expansiveness of the Array API while preserving opportunities for fusion and whole workflow optimization, we pursued a lazy evaluation strategy mediated by a high-level query language.
%
This is implemented by 1) Finch Logic, a minimal, high-level language for expressing array operations and 2) the Physical Interpreter, which executes Finch Logic as a sequence of Finch programs.

### 2024-04-06 sections/eval/morphology.tex
spelling

Some image processing pipelines stand to benefit from structured data processing \cite{donenfeld_unified_2022}.
%
We focus on binary image morpology and the logical transformation of binary images and masks.
%
We consider two operations: binary erosion (computing a mask), and a masked histogram (using a mask to avoid work).
%
We use images are all binary, either by design or having been thresholded.

@@

We compared against OpenCV.
%
We used four datasets. We randomly selected 100 images from the MNIST \cite{lecun_gradient-based_1998} and Omniglot \cite{lake_human-level_2015} character recognition datasets, as well as a dataset of human line drawings \cite{eitz_how_2012}. 
%
We also hand-selected a subset of mask images (these images were less homogenous, so we listed them in Appendix C) from a digital image processing textbook \cite{gonzalez_digital_2006}. 
%
All images were thresholded, and we also include versions of the images that have been magnified before thresholding, to induce larger constant regions. 
%
In our erosion task, the sparseRLE format performs the best as it is asymptotically faster and uses less memory, leading to a 19.5X speedup over OpenCV on the sketches dataset, which becomes arbitrarily large as we magnify the images (here shown as 266X). 
%
We believe the 51.6x on MNIST is due to calling overhead in OpenCV. 
%
The bitwise kernels were effective as well, and would be more effective on datasets with less structure. 
%
A strength of Finch is that it supports structured datasets, even over bitwise operations, allowing us to implement the bitwise kernel and then mask it.

>>>>

Some image processing pipelines stand to benefit from structured data processing \cite{donenfeld_unified_2022}.
%
We focus on binary image morphology and the logical transformation of binary images and masks.
%
We consider two operations: binary erosion (computing a mask), and a masked histogram (using a mask to avoid work).
%
We use images are all binary, either by design or having been thresholded.

@@

We compared against OpenCV.
%
We used four datasets. We randomly selected 100 images from the MNIST \cite{lecun_gradient-based_1998} and Omniglot \cite{lake_human-level_2015} character recognition datasets, as well as a dataset of human line drawings \cite{eitz_how_2012}. 
%
We also hand-selected a subset of mask images (these images were less homogeneous, so we listed them in Appendix C) from a digital image processing textbook \cite{gonzalez_digital_2006}. 
%
All images were thresholded, and we also include versions of the images that have been magnified before thresholding, to induce larger constant regions. 
%
In our erosion task, the SparseRLE format performs the best as it is asymptotically faster and uses less memory, leading to a 19.5X speedup over OpenCV on the sketches dataset, which becomes arbitrarily large as we magnify the images (here shown as 266X). 
%
We believe the 51.6x on MNIST is due to calling overhead in OpenCV. 
%
The bitwise kernels were effective as well, and would be more effective on datasets with less structure. 
%
A strength of Finch is that it supports structured datasets, even over bitwise operations, allowing us to implement the bitwise kernel and then mask it.

### 2024-04-06 sections/eval/spgemm.tex
fixes
spelling

\subsection{Sparse-Sparse Matrix Multiply (SpGEMM)}
\begin{figure}
    \begin{minipage}{0.333\linewidth}
    \begin{minted}{julia}
    @finch begin
      C .= 0
      for j=_
        for i=_
          for k=_
            C[i, j] += AT[k, i] * B[k, j]
          end
        end
      end
      return C
    end
    \end{minted}
    \end{minipage}%
    \begin{minipage}{0.333\linewidth}
    \begin{minted}{julia}
    w = Tensor(SparseByteMap(Element(0)))
    @finch begin
      C .= 0
      for j=_
        w .= 0
        for k=_
          for i=_
            w[i] += A[i, k] * B[k, j]
          end
        end
        for i=_
          C[i, j] = w[i]
        end
      end
    end
    \end{minted}
    \end{minipage}%
    \begin{minipage}{0.333\linewidth}
    \begin{minted}{julia}
    w = Tensor(SparseHash(SparseHash(Element(0))))
    @finch begin
      w .= 0
      for k=_
        for j=_
          for i=_
            w[i, j] += A[i, k] * BT[j, k]
          end
        end
      end
      C .= 0
      for j=_, i=_
        C[i, j] = w[i, j]
      end
    end
    \end{minted}
    \end{minipage}
    \vspace{-12pt}
    \caption{Inner Products, Gustavsons, and Outer Products matrix multiply in Finch}\label{fig:spgemm_listing}
    \vspace{-12pt}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=0.5\linewidth]{spgemm_small_speedup_log_scale.png}%
	\includegraphics[width=0.5\linewidth]{spgemm_joel_speedup.png}
    \vspace{-8pt}
    \caption{A comparison of several matrix multiply algorithms between
    Finch and Taco. On left, we use smaller matrices, ordered from small to big
    dimension. Note that inner-products necessarily requires $O(MN)$ work and
    TACO's outer-products format is dense. Finch can use a sparse outer products
    format and thus has an asymptotic advantage that becomes evident as the
    output dimensions grow. On right, we use only gustavson's algorithm and
    compare on larger matrices.}
    \label{fig:spgemm}
    \vspace{-12pt}
\end{figure}

>>>>

\subsection{Sparse-Sparse Matrix Multiply (SpGEMM)}

\begin{figure}[b]
    \begin{minipage}[t]{0.32\linewidth}
      \vspace{0pt}
    \begin{minted}{julia}
    @finch begin
      C .= 0
      for j=_
        for i=_
          for k=_
            C[i, j] += AT[k, i] * B[k, j]
          end
        end
      end
      return C
    end
    \end{minted}
    \end{minipage}%
    \begin{minipage}[t]{0.33\linewidth}
      \vspace{0pt}
    \begin{minted}{julia}
    w = Tensor(SparseByteMap(Element(0)))
    @finch begin
      C .= 0
      for j=_
        w .= 0
        for k=_
          for i=_
            w[i] += A[i, k] * B[k, j]
          end
        end
        for i=_
          C[i, j] = w[i]
        end
      end
    end
    \end{minted}
    \end{minipage}%
    \begin{minipage}[t]{0.35\linewidth}
      \vspace{0pt}
    \begin{minted}{julia}
    w = Tensor(SparseHash(SparseHash(Element(0))))
    @finch begin
      w .= 0
      for k=_
        for j=_
          for i=_
            w[i, j] += A[i, k] * BT[j, k]
          end
        end
      end
      C .= 0
      for j=_, i=_
        C[i, j] = w[i, j]
      end
    end
    \end{minted}
    \end{minipage}
    \caption{Inner Products, Gustavson's, and Outer Products matrix multiply in Finch}\label{fig:spgemm_listing}
\end{figure}

\begin{figure}[b]
	\includegraphics[width=0.5\linewidth]{spgemm_small_speedup_log_scale.png}%
	\includegraphics[width=0.5\linewidth]{spgemm_joel_speedup.png}
    \vspace{-8pt}
    \caption{A comparison of several matrix multiply algorithms between
    Finch and Taco. On left, we use smaller matrices, ordered from small to big
    dimension. Note that inner-products necessarily requires $O(MN)$ work and
    TACO's outer-products format is dense. Finch can use a sparse outer products
    format and thus has an asymptotic advantage that becomes evident as the
    output dimensions grow. On right, we use only Gustavson's algorithm and
    compare on larger matrices.}
    \label{fig:spgemm}
    \vspace{-12pt}
\end{figure}

@@
